Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-10 paper.tex
Update the intro

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.
Three classes of ARG inference method have emerged as the focus of recent
work, and have all been successfully applied in different contexts. 
Parsimony
based methods~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022}
seek to minimise some aspect of the inferred ARG
(usually the number number of recombinations required), and work well for 
answering specific detailed questions about 
recombination~\citep[e.g.][]{ignatieva_ongoing_2022} when the expected
number of ancestral recombinations is small (tens to hundreds).
The second class of inference method is based on sampling ARGs statistically 
from a population genetics model~\citep{rasmussen_genome-wide_2014,
hubisz_mapping_2020,mahmoudi_bayesian_2022, deng_robust_2024}, and have 
been applied in several
contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. These methods are well-suited to applications 
in which population genetics models are a good fit [TODO: Say this differently, 
I just want to distinguish from e.g. SARS-CoV-2 where pop gen models are 
basically meaningless]
in cases in which the expected number of recombinations is in the 
thousands to tens-of-thousands range (e.g., human datasets in which
we have X samples and a segment of Y Mb of chromosome [TODO quantify
from SINGER paper). 
The third class of inference method and based on heuristics
and approximate models and focus on large scale inference,
where the expected number of recombinations is in the millions or
tens of millions [TODO check numbers here]~\citep{kelleher_inferring_2019,
speidel_method_2019,zhang_biobank-scale_2023}.
Other methods specialised to particular applications have also 
recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021}.
See~\cite{wong_general_2023} for a detailed review of ARG inference 
methods.

@@

Perhaps the clearest route to statistical rigor is through scalable
model-based inference, a central element of population genetics.
The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
tajima_evolutionary_1983} has proved to be a useful model, and is the
basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
The incorporation
of recombination in the coalescent process~\citep{hudson_properties_1983}
has proved to be much more difficult.
% FIXME rephrase, next two sentences, lifted from ARG paper
Early work on inference under the coalescent with recombination
focused on the problem of
inferring the parameters of the
stochastic process, where the ancestry was regarded as a
latent parameter to be averaged out
\citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
nielsen_estimation_2000,
fearnhead_estimating_2001}.
% TODO reuse some of the ideas from the SMC paper here, it has
% some good explanations.
These methods met with limited success
because the state space of ARGs is overwhelmingly large, and
lacks a simple geometry or neighbourhood structure for inference or
sampling methods to  exploit.
The introduction of the Sequentially Markovian Coalescent (SMC)
approximation to the coalescent with recombination
\citep{mcvean_approximating_2005} 
and its refinement in the SMC'~\citep{marjoram_fast_2006},
are the basis of many powerful inference 
% TODO add some more recent stuff
methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
In particular, the SMC is the basis for ARG sampling methods
like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
SINGER~\citep{deng_robust_2024}, which are the gold-standard 
in terms of inference accuracy.
ARGweaver has consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
and, while SINGER has not been evaluated as thoroughly yet, it promises
to be even more accurate as well as substantially more scalable 
computationally~\citep{deng_robust_2024}.

>>>>

\section{Introduction}
Following recent breakthroughs in simulation and inference methods,
there is now intense interest in applying Ancestral Recombination Graphs (ARGs)
to a variety of questions in evolutionary
biology~\citep{lewanski_era_2024,brandt_promise_2024}.
ARGs describe the intricately linked paths
of genetic inheritance resulting from
recombination~\citep{hudson_properties_1983,griffiths_ancestral_1996,wong_general_2023},
and contain rich detail about ancestral processes.
% TODO rephrase this a bit, would be a lot of duplication of references
% in the current formulation.
Numerous applications taking advantage of this detailed history
in inferred ARGs are now appearing~\citep{
stern_approximate_2019,
osmond_estimating_2021,
fan_genealogical_2022,
hejase_deep_2022,
guo_recombination-aware_2022,
ignatieva_ongoing_2022,
wang_complex_2022,
zhang_biobank-scale_2023,
nowbandegani_extremely_2023,
ignatieva_distribution_2023,
fan_likelihood_2023,
huang_estimating_2024,
grundler_geographic_2024,
korfmann_simultaneous_2024,
deraje_inferring_2024}
and it seems likely that many more will
follow~\citep{harris_database_2019,harris_using_2023}.

Three classes of ARG inference method have emerged as the focus of recent
work, and have all been successfully applied in different contexts. 
Parsimony
based methods~\citep{ignatieva_kwarg_2021,rasmussen_espalier_2022}
seek to minimise some aspect of the inferred ARG
(usually the number number of recombinations required), and work well for 
answering specific detailed questions about 
recombination~\citep[e.g.][]{ignatieva_ongoing_2022} when the expected
number of ancestral recombinations is small (tens to hundreds).
The second class of inference method is based on sampling ARGs statistically 
from a population genetics model~\citep{rasmussen_genome-wide_2014,
hubisz_mapping_2020,mahmoudi_bayesian_2022, deng_robust_2024}, and have 
been applied in several
contexts~\citep[e.g.][]{de_chimpanzee_2016,shriner_whole_2018,hejase_genomic_2020,
stankowski_genetic_2024}. These methods are well-suited to applications 
in which population genetics models are a good fit [TODO: Say this differently, 
I just want to distinguish from e.g. SARS-CoV-2 where pop gen models are 
basically meaningless]
in cases in which the expected number of recombinations is in the 
thousands to tens-of-thousands range (e.g., human datasets in which
we have X samples and a segment of Y Mb of chromosome [TODO quantify
from SINGER paper). 
The third class of inference method is based on heuristics
and approximate models and focus on large scale inference,
where the expected number of recombinations is in the millions or
tens of millions [TODO check numbers here]~\citep{kelleher_inferring_2019,
speidel_method_2019,zhang_biobank-scale_2023}.
Other methods specialised to particular applications have also 
recently been proposed~\citep{zhan_towards_2023,schaefer_ancestral_2021};
see~\cite{wong_general_2023} for a detailed review of ARG inference 
methods.

Statistically rigorous ARG sampling methods like
ARGweaver~\citep{rasmussen_genome-wide_2014} and
SINGER~\citep{deng_robust_2024} have many advantages. 
Firstly, ARGweaver has consistently out-performed other methods in terms 
of accuracy in a variety of benchmarks~\citep{brandt2022evaluation}[MORE
citations,anywhere ARGweaver has been tested for accuracy]
and, while SINGER has not been evaluated as thoroughly yet, it promises
to be even more accurate~\citep{deng_robust_2024}. 
Secondly, the direct connection with
population genetic models greatly aids interpretability, and allows for
natural extensions like the joint estimation of ARGs and
demography~\citep{hubisz_mapping_2020}. Thirdly, sampling ARGs from a
posterior distribution provides a means of quantifying the uncertainty 
around estimates, which can be used in downstream applications [citation
of this?]. Heuristic, approximate inference methods like
Relate~\citep{speidel_method_2019}, tsinfer~\citep{kelleher_inferring_2019}
and ARG-Needle~\citep{zhang_biobank-scale_2023} on the other hand 
tend to be less accurate

@@

% Perhaps the clearest route to statistical rigor is through scalable
% model-based inference, a central element of population genetics.
% The coalescent~\citep{kingman_genealogy_1982,kingman_coalescent_1982,hudson_testing_1983,
% tajima_evolutionary_1983} has proved to be a useful model, and is the
% basis of a wide variety of inferences [some citations, e.g Hey IM model]. 
% The incorporation
% of recombination in the coalescent process~\citep{hudson_properties_1983}
% has proved to be much more difficult.
% % FIXME rephrase, next two sentences, lifted from ARG paper
% Early work on inference under the coalescent with recombination
% focused on the problem of
% inferring the parameters of the
% stochastic process, where the ancestry was regarded as a
% latent parameter to be averaged out
% \citep[e.g.][]{griffiths_ancestral_1996,kuhner_maximum_2000,
% nielsen_estimation_2000,
% fearnhead_estimating_2001}.
% % TODO reuse some of the ideas from the SMC paper here, it has
% % some good explanations.
% These methods met with limited success
% because the state space of ARGs is overwhelmingly large, and
% lacks a simple geometry or neighbourhood structure for inference or
% sampling methods to  exploit.
% The introduction of the Sequentially Markovian Coalescent (SMC)
% approximation to the coalescent with recombination
% \citep{mcvean_approximating_2005} 
% and its refinement in the SMC'~\citep{marjoram_fast_2006},
% are the basis of many powerful inference 
% % TODO add some more recent stuff
% methods~\citep[e.g.]{li_inference_2011,paul_accurate_2011,schiffels_inferring_2014}.
% In particular, the SMC is the basis for ARG sampling methods
% like ARGweaver~\citep{rasmussen_genome-wide_2014} and 
% SINGER~\citep{deng_robust_2024}, which are the gold-standard 
% in terms of inference accuracy.
% ARGweaver has consistently out-performed other methods in terms 
% of accuracy in a variety of benchmarks~\citep{brandt2022evaluation},
% and, while SINGER has not been evaluated as thoroughly yet, it promises
% to be even more accurate as well as substantially more scalable 
% computationally~\citep{deng_robust_2024}.
